Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-07 manu-draft.qmd
into draft almost done but needs work

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ in areal cover since the 1980s [@waycott2009]. These losses are comparable, if not more, than other critical coastal environments, which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require an adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline.

The sustained coverage or restoration of seagrasses by means other than planting requires management actions that create suitable conditions supportive for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses. There are limited examples of successful recovery of seagrass through control of nutrient inputs, with notable exceptions including Tampa Bay\... examples in Orth 2023. Lack of recovery of seagrasses through nutrient controls alone suggests insufficient reductions of nutrient inputs to promote restoration, such as for ecosystems where hysteresis may substantially limit the reverse of a trajectory deline. For coastal systems More likely, the role of alternative stressors may also be inadequately considered and addressed in management paradigms.

Climate change has complicated the understanding of ecosystem response to conventational stressors, presenting new challenges and expectations for how ecological resources will respond to management actions. rising temperature

el nino effects, see Greening and Janicki 2006

Heat stress combined with highlight accelerates decline of *E. acoroides*, used 36C [@zhang2023]

Widgeongrass colonization in Chesapeake Bay as an opportunistic, heat-tolerant species that has replaced Eelgrass, although former is sensitive to nutrient pulses. Paper provides an example of implications on climate change and nutrient management on foundation species and system resiliency [@hensel2023].

This study focuses on Tampa Bay...

Background: SST trends in Tampa Bay w/ EPC data, WFS, and deep GOM [@nickerson2023] showed that 1975 to 2022 trend for EPC was notable, but less so from 1998 to 2022. Also noted the trend was most pronounced in the winter, see Fig 3

Good CC references for FL: https://www.mdpi.com/2071-1050/15/14/11364

@lewis1985 review of seagrass in Tampa Bay. @lirman2003 conducted exposure experiments to evaluate seagrass growth in response to a range of salinity conditions. @mcmillan1967 discusses growth of halodule, syringodium, thalassia, and ruppia in response to salinity increases (up to 75 psu), no info on lower limit. Cites @phillips1960 for a salinity range of Thalassia in Florida of 33 - 38 psu. @zieman1975 discusses seasonal variation of thalassia relative to temp and salinity

Temp background for TB: NOAA coastwatch trends

>>>>

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ in areal cover since the 1980s [@waycott2009]. These losses are comparable, if not more, than other critical coastal environments, which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create suitable conditions supportive for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls. In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis where the path to recovery is not the same as the path to decline. Costly alternative management actions in addition to nutrient reductions are needed in these situations. Management actions are much more effective in estuarine sytems that are stressed but not past critical thresholds that define an alternative state.

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns. These changes will profoundly alter physical habitats, creating suboptimal or uninhabitable conditions for many species. @lefcheck2017 evaluated interactions between water clarity and rising temperatures with over 30 years of data in Chesapeake Bay. The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors. Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant eelgrass (*Zostera marina*) given higher tolerance to increased temperatures [@hensel2023]. Changing frequency and severity of precipicitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023] , suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years. Through successful reduction of nitrogen loads by 2/3 from , seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023], whereas changing salinity has not been well-described, nor have potential links to seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.
